Exploring probability distributions for bivariate temporal granularities

### 2019-11-05 main.Rmd
visualization section rough draft update

Before exploring the behavior of a "dependent variable" across time granularities,
it is important to know how these granularities interact with each other. If two time granulaties of interest interact, the relationship between each of the interacting variables and the dependent variable will depend on the value of the other interacting variable, For illustration, distribution of half-hourly electricity consumption of Victoria is plotted across different time granularities in each of the panel in
\autoref{fig:allFig}. \autoref{fig:allFig} (a) shows the letter value plot across days of the month faceted by months like January, March and December. \autoref{fig:allFig} (c) shows box plot  across days of the year by the 1st, 15th, 29th and 31st days of the month. \autoref{fig:allFig} (d) showing violin plot across days of the month faceted by week of the month. \autoref{fig:allFig} (e), variations across week of the year conditional on week of the month can be observed through a ridge plot and \autoref{fig:allFig} (f) shows decile plots across  day of the year and month of the year.

@@

```{r allFig, fig.height=5.5, out.width="50%", fig.pos = "p", fig.align= 'left',echo=FALSE, eval=TRUE, warning=FALSE,message = FALSE, fig.cap="Various probability distribution plots of electricity consumption data of Victoria from 2012 to 2014. (a) Letter value plot by DoM and MoY, (b) Decile plot by HoD and DoW (c) Box plot by DoY and DoM, (d) Violin plot of DoM and WoM, (e) Ridge plot by WoM and WoY, (f) Decile plot by DoY and MoY. Only plots (a) and (b) show harmonised time variables.", fig.show = 'hold'}

To define a general framework, we define harmony and clashes. Harmonies are the pair of granularities that aid the process of exploratory analysis and clashes are pair that interact in a non-compatible ways and hinder the process of exploration of the dependent variable if plotted together.

@@

<!-- Why visualize -->
Analysts often want to fit their data to statistical models, either to test hypotheses or predict future values. However, improper choice of models can lead to wrong predictions. One important use of visualization is exploratory data analysis, which is gaining insight into how data is distributed to inform data transformation and modeling decisions.

@@

<!-- Why distribution plot -->
<!-- Huge data difficult to summaries using one statistic-->
But with huge amount of data being available, sometimes mean, median or any one summary statistic is not enough to understand a data set. Soon enough following questions become more interesting:

@@

This is when displaying a probability distribution becomes a potentially useful approach.

<!-- What all can you see from a distribution plot-->

The entire distribution can be visualized or some contextual summary statistics can be visualized to emphasize certain properties of the distribution. These properties can throw light on central tendency, skewness, kurtosis, variation of the distribution and can also be useful in detecting extreme behavior or anomalies in the data set.

@@

The time series variable can be plotted against many time granularities to get more understanding of the underlying periodicty, however, we will restrict ourselves to see the distribution of the time series across bivariate temporal granularities. That neccessiates plotting one temporal granularity along the x-axis and the other one across facets.

Now, due to the hierarchical arrangement of the granularities, there are certain granularities which when plotted together do not give us the layout to do exploration, for example, structurally empty combinations (clashes) are not recommended to plot together. The harmonies when plotted together can help exploration. But still the question remains that which distribution plot should be chosen to bring out the best of exploratory data analysis. This is a function of which features of the distribution we are interested to look at, how much display space is available to us and also if the number of observations are enough for that distribution plot.

@@

Recommendations for distribution plots depend on the levels(very high/high/medium/low) of the two granularities plotted. They will vary depending on which granularity is placed on the x-axis and which one across facets. Assumptions are made to ensure display is not too cluttered by the space occupied by various kinds of distribution plots. Moreover, the recommendation system ensures that there are just enough observations before choosing a distribution plot.

Levels are categorized as very high/high/medium/low each for the facet variable and the x-axis variable. Default values for these levels are based on levels of common temporal granularities like day of the month, day of a fortnight or day of a week. For example, any levels above 31 is considered as very high, any levels between 14 to 31 are taken as high and that between 7 to 14 is taken as medium and below 7 is low. 31, 14 and 7 are the levels of days-of-month, days-of-fortnight and days-of week respectively. These default values are decided based on usual cognitive power while comparing across facets and display size available to us. Let us consider case by case and see which plots are better suitable in which scenarios.

<!-- Graphical output devices are flat. If there are only two variables, a cartesian graph of one variable against the other can show the configuration of points. As soon as the data move to three variables, we must be content in attempting to infer multi-dimensional structure by a two-dimensional medium.  -->

Facetting creates small multiples of data each representing a different subset of the data. Small multiples are powerful for exploratory data analysis where we can rapidly compare patterns in different parts of the data and see whether they are the same or different. But space and resolution quickly become a problem if the number of levels of facet variables is too high. Also, Cleveland suggested that a graphic should display as much information as it can, with the lowest possible cognitive strain to the viewer.

@@

@@

@@

Suppose we have $T$ hourly observations, and $C_1$ is HOD and $C_2$ is DOW. Each element of $C_1$ occurs approximately $T/n_1$ times while each element of $C_2$ occurs approximately $T/n_2$ times. There are no structurally empty combinations, and each combination will occur on average an equal number of times as $T\rightarrow\infty$, so the average number of observations per combination is $T/(n_1n_2)$. If we require at least $k$ observations to create a meaningful panel, then provided $T \ge n_1n_2k$, the visualization will be acceptable. The value of $k$ will depend on what type of visualization we are producing. For scatterplots, even $k=3$ may be acceptable, but for density estimates, we probably need $k \ge 30$.

The above discussion shows that problems will occur when $n_1n_2 > T/k$, even without structurally empty combinations.

Problems will also occur when we have rarely occurring categories such as the 366th day of the year, or the 31st day of the month.

### Homogeneity within and across facets

We should be cautious to plot two granularities which vary largely across or within facets. Even if for each combinations, the number is good enough for plotting distribution, large differences between number of observations between and across facets, might have an impact on the emperical dfistribution. We use Gini's measure to compute if differences in number of observations across or within facets are significantly different.

<!-- Lit review on statistical distribution plot-->
 Statistical distribution plots

Most commonly used techniques to display distribution of data include the histogram (Karl Pearson), which shows the prevalence of values grouped into bins and the box-and-whisker plots [@Tukey1977-jx] which convey statistical features such as the median, quartile boundaries, hinges, whiskers and extreme outliers. The box plot is a compact distributional summary, displaying less detail than a histogram. Due to wide spread popularity and simplicity in implementation, a number of variations are proposed to the original one which provides alternate definitions of quantiles, whiskers, fences and outliers. Notched box plots [@Mcgill1978-hg, 1978] has box widths proportional to the number of points in the group and display confidence interval around medians aims to overcome some drawbacks of box plots.

The vase plot [@Benjamini1988-io, 1988] was a major revision from the concept of box plots where the width of box at each point is proportional to estimated density. Violin plots [@Hintze1998-zi, 1998] display the density for all data points and not only the box. The summary plot [@Potter2010-qc, 2010] combines a minimal box plot with glyphs representing the first five moments (mean, standard deviation, skewness, kurtosis and tailings), and a sectioned density plot crossed with a violin plot (both color and width are mapped to estimated density), and an overlay of a reference distribution. The highest density region (HDR) box plot proposed by [@Hyndman1996-ft] displays a probability density region that contains points of relatively highest density. The probabilities for which the summarization is required can be chosen based on the requirement. These regions do not need to be contiguous and help identify multi-modality. The letter-value box plot [@Hofmann2017-sg, 2006] was designed to adjust for number of outliers proportional to the data size and display more reliable estimates of tail. Because this display just adds extra letter values, it suffers from the same problems as the original box plot, and multimodality is almost impossible to spot[@Wickham_undated-vr].

Moreover, much like the quartiles divide the data set equally into four equal parts, extensions might include dividing the data set even further. The deciles plots consist of 9 values that split the data set into ten parts and the percentile plot consists of 99 values that split the data set into hundred parts. A large data set is required before the extreme percentiles can be estimated with any accuracy.

Finally, a density plot which uses a kernel density estimate to show the probability density function of the variable can show the entire distribution. Also, a Ridge line plot (sometimes called Joy plot) shows the distribution of a numeric value for several groups. Distribution can be represented using histograms or density plots, all aligned to the same horizontal scale and presented with a slight overlap.

Recommendation system that aid exploration

>>>>

@@

Before exploring the behavior of a "dependent variable" across time granularities,
it is important to know how these granularities interact with each other. If two time granulaties of interest interact, the relationship between each of the interacting variables and the dependent variable will depend on the value of the other interacting variable. To define a general framework, we define harmony and clashes. Harmonies are the pair of granularities that aid the process of exploratory analysis and clashes are pair that interact in a non-compatible ways and hinder the process of exploration of the dependent variable if plotted together.

@@

<!-- Why visualize -->
<!-- Analysts often want to fit their data to statistical models, either to test hypotheses or predict future values. However, improper choice of models can lead to wrong predictions. One important use of visualization is exploratory data analysis, which is gaining insight into how data is distributed to inform data transformation and modeling decisions. -->

@@

<!-- Why distribution plot -->
<!-- Huge data difficult to summaries using one statistic-->
With huge amount of data being available, mean, median or any one summary statistic might not enough to understand a data set. Soon enough following questions become more interesting:

@@

This is when displaying a probability distribution becomes a potentially useful approach.
The entire distribution can be visualized or some contextual summary statistics can be visualized to emphasize certain properties of the distribution. These properties can throw light on central tendency, skewness, kurtosis, variation of the distribution and can also be useful in detecting extreme behavior or anomalies in the data set.

@@

<!-- Why 2D -->
Graphical output devices are flat. If there are only two variables, a cartesian graph of one variable against the other can show the configuration of points. As soon as the data move to three variables, we need to infer multi-dimensional structure by a two-dimensional medium[Cleveland, elements of graphing data]. Any attempt to encode all these temporal granularities together to develop insights on periodicity might fail or become clumsy. Instead, the big problem can be broken down into smaller pieces by focusing on specific tasks per visual representation in a 2D space.

<!-- Why facet-ing -->
`ggplot2` (Wickham, 2016) is an excellent tool that facilitates the process of mapping different variables to a 2D frame through grammar of graphics. One aspect of ggplot2 is faceting, inspired by Trellis plots (Cleveland, 1993). These are based on conditioning the values taken on by one or more of the variables in a data set. This means creating same plots using data subsets corresponding to each level of the conditioned variable. This is a powerful tool for exploratory data analysis as we can rapidly compare patterns in different parts of the data.

<!-- What are we visualizing -->
We will see the distribution of the time series across bivariate temporal granularities through faceting approach with one temporal granularity plotted along the x-axis, the other one across facets and the dependent time series variable on the y-axis.

Now, different distribution plots might be appropriate depending on which features of the distrbution we are interested to look at, the levels of time granularities being plotted, how granularities interact and number of observations available. We will look at each of these aspects separately and analyze the effect of each on visualization.

<!-- Now, due to the hierarchical arrangement of the granularities, there are certain granularities which when plotted together do not give us the layout to do exploration, for example, structurally empty combinations (clashes) are not recommended to plot together. The harmonies when plotted together can help exploration. But still the question remains that which distribution plot should be chosen to bring out the best of exploratory data analysis. This is a function of which features of the distribution we are interested to look at, how much display space is available to us and also if the number of observations are enough for that distribution plot. -->

@@

<!-- Lit review on statistical distribution plot-->

<!-- Description and lit review of boxplot -->
There are several ways to plot statistical distributions. The displayed probabilities in these plots are either computed using kernel density estimation methods or emperical statistical methods. We will discuss the benefits and challenges of few conventional and recent ways to plot distributions using both of these methods.

### Emperical methods

Most commonly used techniques to display distribution of data include the histogram (Karl Pearson), which shows the prevalence of values grouped into bins and the box-and-whisker plots [@Tukey1977-jx] which convey statistical features such as the median, quartile boundaries, hinges, whiskers and extreme outliers. The box plot is a compact distributional summary, displaying less detail than a histogram. Due to wide spread popularity and simplicity in implementation, a number of variations are proposed to the original one which provides alternate definitions of quantiles, whiskers, fences and outliers. Notched box plots [@Mcgill1978-hg, 1978] has box widths proportional to the number of points in the group and display confidence interval around medians aims to overcome some drawbacks of box plots.<!-- Challenge of boxplot --> The standard box plot and all of these variations are helpful to get an idea of the distribution at a glance. However, we do not have an idea if the distribution is multimodal. Moreover, for data less than 1000 observations, detailed estimates of tail behavior beyond the quartiles are not trustworthy. Also, the number of outliers is large for larger data set since number of outliers is proportional to the number of observations.

<!-- Description and lit review of letter-value plots -->
The letter-value box plot [@Hofmann2017-sg, 2006] was designed to adjust for number of outliers proportional to the data size and display more reliable estimates of tail.It shows only actual data values and no smoothed summaries. However, the letter values are shown till the depths where the letter values are reliable estimates of their corresponding quantiles. These might lead to a lot of letter values being shown and leading to overload of information in one plot. Also, multimodality is difficult to spot.

<!-- Description and lit review of quantile plots -->

Quantile plots visually portray the quantiles of the distribution of sample data. Much like the quartiles divide the data set equally into four equal parts, extensions might include dividing the data set even further. These plots are referred to as "quantile" plots with, where number of quantiles 9 for deciles and 99 for percentiles. No distributional assumptions are made about the data since empirical deciles are plotted. It avoids much clutter and just enable us to focus on specific probabilities. These plots do not allow us to see the shape, skewness or nature of the tail of the distribution with so much clarity as violin plots, ridge plots or letter value plots. Also, a large data set is required for the extreme percentiles to be estimated with any accuracy.

<!-- Description and lit review of density plots -->

<!-- The vase plot [@Benjamini1988-io, 1988] was a major revision from the concept of box plots where the width of box at each point is proportional to estimated density.  -->

### Kernel density estimation methods

A density plot which uses a kernel density estimate to show the probability density function of the variable can show the entire distribution, unlike discretizing the distribution and showing only parts of it.

Violin plots [@Hintze1998-zi, 1998] adds the information available from local density estimates to summary statistics provided by box plots. The shape of the violin represents the density estimate of the variable. The more data points in a specific range, the larger the violin is for that range. Adding two density plots gives a symmetric plot which makes it easier to see the magnitude of the density and compare across categories. Howevr, the density in violin are estimated through kernel density estimation and thus makes assumptions when selecting kernel or bandwidth. As a result, the plot shows smooth summaries based on the assumptions and not only on actual observations.

The summary plot [@Potter2010-qc, 2010] combines a minimal box plot with glyphs representing the first five moments (mean, standard deviation, skewness, kurtosis and tailings), and a sectioned density plot crossed with a violin plot (both color and width are mapped to estimated density), and an overlay of a reference distribution. This suffers from the same problem as boxplots or violin plot, as it is combination of those two.

<!-- Description and lit review of density plots -->
 A Ridge line plot (sometimes called Joy plot) shows the distribution of a numeric value for several groups. Distribution can be represented using histograms or density plots, all aligned to the same horizontal scale and presented with a slight overlap. A clear advantage over boxplots is that these plots allow us to see multimodality in the distribution. However, these plots can be obscuring when there is overlap of distribution for two or more categories of the y-axis. Also, if there are lot of categories, it is difficult to compare the height of the densities across categories.

<!-- Description and lit review of HDR plots -->
The highest density region (HDR) box plot proposed by [@Hyndman1996-ft] displays a probability density region that contains points of relatively highest density. The probabilities for which the summarization is required can be chosen based on the requirement. These regions do not need to be contiguous and help identify multi-modality.

As a general rule, the quantile plots are really useful for picking patterns, whereas, other more involved methods of plotting are useful for studying anomalies or outlier behavior. It is good to be aware of the benefits and challenges while choosing any distribution plots, depending on the context.

### Levels of temporal granularities

Choice of plots depend on the levels(very high/high/medium/low) of the two granularities plotted. The criteria for different levels could be based on usual cognitive power while comparing across facets and display size available to us.Space and resolution become a problem if the number of levels of facet variables is too high. Hence, plot choices will also vary depending on which granularity is placed on the x-axis and which one across facets.

Levels are categorized as very high/high/medium/low each for the facet variable and the x-axis variable. Default values for these levels are chosen based on levels of common temporal granularities like day of the month, day of a fortnight or day of a week. Any levels above 31 can be considered as very high, any levels between 14 to 31 can be taken as high and that between 7 to 14 can be taken as medium and below 7 as low. 31, 14 and 7 are the levels of days-of-month, days-of-fortnight and days-of week respectively.

Space and cognitive power both are constrained if the number of levels of facet variables is very high. So as a general rule, we should avoid plotting a temporal granularity with very high levels across facets.

 - If levels of both the granularities are low, then any distribution plots might be chosen depending on which feature of the distribution who is interested to look at.

 - If level of the granularity plotted across x-axis is more than medium, ridge plots should be avoided to escape overlap of categories.

 - If level of the granularity plotted across x-axis is low, then any distribution plots might be chosen depending on which feature of the distribution who is interested to look at, irrespective of the number of the levels of the facet variable.

 - If level of the granularity plotted across x-axis is more than or equal to high, quantile plots are preferred.

<!-- Let us consider case by case and see which plots are better suitable in which scenarios. -->

<!--  Assumptions should be made to ensure display is not too cluttered by the space occupied by various kinds of distribution plots.  -->
<!-- Moreover, the recommendation system ensures that there are just enough observations before choosing a distribution plot. -->

<!-- Levels are categorized as very high/high/medium/low each for the facet variable and the x-axis variable. Default values for these levels are based on levels of common temporal granularities like day of the month, day of a fortnight or day of a week. For example, any levels above 31 is considered as very high, any levels between 14 to 31 are taken as high and that between 7 to 14 is taken as medium and below 7 is low. 31, 14 and 7 are the levels of days-of-month, days-of-fortnight and days-of week respectively. These default values are decided based on usual cognitive power while comparing across facets and display size available to us. Let us consider case by case and see which plots are better suitable in which scenarios. -->

<!-- Facetting creates small multiples of data each representing a different subset of the data. Small multiples are powerful for exploratory data analysis where we can rapidly compare patterns in different parts of the data and see whether they are the same or different. But space and resolution quickly become a problem if the number of levels of facet variables is too high. Also, Cleveland suggested that a graphic should display as much information as it can, with the lowest possible cognitive strain to the viewer. -->

@@

For illustration, distribution of half-hourly electricity consumption of Victoria is plotted across different time granularities in each of the panel in
\autoref{fig:allFig}. \autoref{fig:allFig} (a) shows the letter value plot across days of the month faceted by months like January, March and December. \autoref{fig:allFig} (c) shows box plot  across days of the year by the 1st, 15th, 29th and 31st days of the month. \autoref{fig:allFig} (d) showing violin plot across days of the month faceted by week of the month. \autoref{fig:allFig} (e), variations across week of the year conditional on week of the month can be observed through a ridge plot and \autoref{fig:allFig} (f) shows decile plots across  day of the year and month of the year.

@@

```{r allFig, fig.height=5.5, out.width="50%", fig.pos = "h", fig.align= 'left',echo=FALSE, eval=TRUE, warning=FALSE,message = FALSE, fig.cap="Various probability distribution plots of electricity consumption data of Victoria from 2012 to 2014. (a) Letter value plot by DoM and MoY, (b) Decile plot by HoD and DoW (c) Box plot by DoY and DoM, (d) Violin plot of DoM and WoM, (e) Ridge plot by WoM and WoY, (f) Decile plot by DoY and MoY. Only plots (a) and (b) show harmonised time variables.", fig.show = 'hold'}

@@

 Suppose we have $T$ hourly observations, and $C_1$ with $n_1$ categories and $C_2$ with $n_1$ categories. Each element of $C_1$ occurs approximately $T/n_1$ times while each element of $C_2$ occurs approximately $T/n_2$ times. There are no structurally empty combinations, and each combination will occur on average an equal number of times as $T\rightarrow\infty$, so the average number of observations per combination is $T/(n_1n_2)$. If we require at least $k$ observations to create a meaningful panel, then provided $T \ge n_1n_2k$, the visualization will be acceptable. The value of $k$ will depend on what type of visualization we are producing. For emperical estimates, even $k=10$ may be acceptable, but for density estimates, we probably need $k \ge 30$.

<!-- The above discussion shows that problems will occur when $n_1n_2 > T/k$, even without structurally empty combinations. -->

Let us take a specific example to illustrate this. Rarely occurring categories such as the 366th day of the year, or the 31st day of the month can suffer from such problem.

### Unevenly distributed events

Even when there are no rarely occuring events, when number of observations are not evenly distributed between different categories of the time granularities.Large differences between number of observations between and across facets, might have an impact on the estimation of dfistribution. We use Gini's measure to compute if differences in number of observations across or within facets are significantly different.
